Reducing RES Droughts through the integration of wind and solar PV

### 2025-04-11 Article/RES_droughts_Morin_ReEn_R3.tex
Reapply "Update RES_droughts_Morin_ReEn_R3.tex"
This reverts commit c23a2f25f530743537f59824f32df1171d5de3ca.

Generic datasets for wind and solar PV CF are often used for the quantification of RES droughts. Despite undergoing a general validation process, they are often not fully representative of each geographical location, and can show differences in the number of RES drought events subsequently identified~\citep{kies2021drought}. This study evaluates the skill of a dataset developed for the European region (C3SE) when applied to a specific country; Ireland. In particular, the analysis explores the impact of using a generic versus a tailored dataset on RES drought assessments, in the context of a transition from a wind-dominated system to one with a greater share of solar PV capacity.

@@

The wind CF in C3SE was calculated using wind speeds at 100 m ($u_{100}$, $v_{100}$) and a standard wind turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. As data on wind turbine fleet locations and specifications are difficult to obtain across Europe, C3SE assumes a homogeneous distribution of wind turbines across the ERA5 grid. While this approach does not capture the precise capacity factors reported by grid operators, it provides a well-correlated time series that effectively represents the impact of climate variability on wind power generation. The C3SE solar PV CF was also calculated for the ERA5 grid. It is derived from meteorological data, including surface solar radiation downwards ($ssrd$) and air temperature ($t2m$), using a reference solar PV plant model. This model incorporates empirical calculations for key system components such as optical losses, module efficiency, and inverters. The final CF accounts for a mix of module orientations typical for each location~\citep{saintdrenan2018solar}.

@@

The three datasets provide CF time series for both wind and solar PV. In addition to analysing the CF of wind and solar PV separately, a combined CF was computed for each dataset by averaging wind and solar PV CF, weighted by their installed capacities at the end of 2023 (5.9 GW for wind power and 0.6 GW for solar PV power). This configuration is referred to as the 91W-9PV scenario, reflecting the distribution of 91\% wind and 9\% solar PV capacity. Given that solar PV capacity in Ireland is low in 2023, and to explore how a more balanced distribution of wind and solar PV capacities might impact RES droughts, this study also considered a second scenario, referred to as 57W-43PV, where the installed solar PV capacity is assumed to increase to 8.6 GW, while wind capacity rises to 11.45 GW. These values are based on targets outlined in the roadmap published by the 2024 Climate Action Plan~\citep{cap2024future}. This study does not include offshore wind in the analysis. Recent reports suggest that even by 2030, Ireland is unlikely to have any significant new offshore wind farms, with projected offshore capacity expected to remain near zero using realistic scenarios~\citep{seai2024future}.

@@

The accuracy of the datasets used in this study was verified, before continuing to the analysis of RES droughts. For the verification process, time-varying values of installed capacity were used to account for changes in RES development over the verification period. This step allowed us to assess how well the datasets represent renewable generation profiles by comparing them against observed data. This validation step evaluates how well the datasets represent actual renewable energy production by comparing them against observed data. The overall statistical distribution of CF values for wind (2014–2023) and solar PV (2023) is presented in the violin plots in Fig.~\ref{fig:violin_plots}. These plots illustrate the density of CF values for each dataset, highlighting their differences and alignment with observations. The results indicate that ATL aligns more closely with observations for wind, while all datasets exhibit similar distributions for solar PV.

@@

Verification of wind generation data highlights the importance of selecting a representative wind turbine power curve for the region being analysed. The ATL dataset, which uses a representative wind turbine power curve, is skilled at reproducing wind CF and RES droughts across Ireland. On the other hand, the power curve used for both C3S~GRD and C3S~NAT is not representative for Ireland, as it severely overestimates generation, underestimating the occurrence of RES droughts. This highlights a problem with using generalised datasets for analysing RES droughts: biases severely affect their ability to accurately reproduce RES drought events. The skill scores for the three datasets (Table~\ref{tab:wind_skill_scores}) show only a small difference in their ability to reproduce the changes in CF, as seen by their similar CC scores. However, their ability to reproduce the actual CF values is much lower than that for ATL, with RMSE scores almost four times higher than for the two C3S datasets. There is a clear bias towards an overestimation of CF, seen in the MBE values, which leads to an underestimation of RES droughts. This highlights the need to use regionally verified datasets to assess RES droughts.

@@

In this section, RES droughts are analysed by calculating the frequency and duration of RES drought events, the return periods for different RES drought durations, and the seasonality of RES drought events. Understanding the characteristics and timing of RES drought events enables system operators to optimally plan for reserve capacity requirements, ensuring grid stability and security of supply. Results are presented for the three datasets, allowing their differences on the characterisation of RES droughts to be clearly identified.

@@

RES drought events identified over the 45-year period were used to calculate the return periods for different RES drought durations. A return period is the estimated average time interval between events of a specified duration (not to be confused with the frequency of their occurrence within a fixed time frame). Fig.~\ref{fig:return_periods} shows the return periods for different RES drought durations, which can be used to capture the most extreme events affecting the system. Understanding their return periods is crucial, as extreme but rare RES droughts pose the toughest challenge to energy security by placing significant strain on the conventional backup sources necessary to maintain security of supply during these events.

@@

Across Fig.~\ref{fig:return_periods}a, c, and d, the return periods in the ATL dataset are consistently higher than those in the two C3S datasets. For instance, in the 91W-9PV scenario (Fig.~\ref{fig:return_periods}c), an event with a one-year return period lasts six days in the ATL dataset, compared to only five days in the C3S datasets. This difference underscores the importance of dataset selection when quantifying RES droughts, as each dataset’s assumptions and parametrisations significantly influence RES droughts duration estimates. Additionally, in all four graphs, the similarity between results from the two C3S datasets suggests that assumptions in the ATL dataset, such as wind turbine power curve selection and PV panel specifications, have a greater impact on RES drought duration estimates than the precise geographic distribution of RES farms when studying the return periods of RES droughts.

@@

The seasonal analysis of RES droughts is based on the percentage of hours in each month classified as part of a RES drought event. Wind droughts tend to be more frequent during summer, whereas solar PV droughts are more common in winter due to reduced sunlight. By comparing these seasonal patterns across different datasets and energy scenarios, this study examines how dataset-specific assumptions and variations in capacity mix affect the overall characterisation of RES drought events.

@@

The effect of the integration of solar PV capacity in a wind-dominated system on RES droughts has been explored. Our analysis has demonstrated that transitioning to a system with more equal amounts of wind and solar PV capacity reduces the occurrence of RES drought events, mitigates extreme RES drought conditions and enhances overall system resilience. This improvement is attributed to the complementary nature of wind and solar PV generation, as solar PV generation typically peaks in summer while wind generation predominates during winter. However, this integration is unable to counter critical winter RES droughts, which coincide with the strongest electricity demand in Northern European countries.

The results presented in this study have three main limitations. First, the definition of RES droughts based on generation does not consider the important role of demand, which could be of interest to system operators. Second, recent solar PV capacity expansions have changed the generation profile, limiting solar PV data for model training to a single year, although a longer validation period would be preferable. Third, the source for weather data is ERA5 has limited spatial resolution, an issue that can be addressed once higher resolution datasets become available.

>>>>

Generic datasets for wind and solar PV CF are often used for the quantification of RES droughts. Despite undergoing a general validation process, they are often not fully representative of each geographical location, and can show differences in the number of RES drought events subsequently identified~\citep{kies2021drought}. This study evaluates the skill of a dataset developed for the European region (C3SE) when applied to a specific country; Ireland, representative of Northwestern Europe. In particular, the analysis explores the impact of using a generic versus a tailored dataset on RES drought assessments, in the context of a transition from a wind-dominated system to one with a greater share of solar PV capacity.

@@

The wind CF in C3SE was calculated using wind speeds at 100 m ($u_{100}$, $v_{100}$) and a standard wind turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. As data on wind turbine fleet locations and specifications are difficult to obtain across Europe, C3SE assumes a homogeneous distribution of wind turbines across the ERA5 grid. While this approach does not capture the precise capacity factors reported by grid operators, it provides a well-correlated time series that effectively represents the impact of weather variability on wind power generation. The C3SE solar PV CF was also calculated for the ERA5 grid. It is derived from meteorological data, including surface solar radiation downwards ($ssrd$) and air temperature ($t2m$), using a reference solar PV plant model.  This model incorporates empirical calculations for key system components, such as optical losses, the electrical characteristics of the power module and the power curve of the PV inverter. The final CF accounts for a mix of module orientations typical for each location~\citep{saintdrenan2018solar}.

@@

The three datasets provide time series for both wind and solar PV CF. In addition to analysing the CF of wind and solar PV separately, a combined CF was computed for each dataset by averaging wind and solar PV CF, weighted by their installed capacities at the end of 2023 (5.9 GW for wind power and 0.6 GW for solar PV power). This configuration is referred to as the 91W-9PV scenario, reflecting the distribution of 91\% wind and 9\% solar PV capacity. Given that solar PV capacity in Ireland is low in 2023, and to explore how a more balanced distribution of wind and solar PV capacities might impact RES droughts, this study also considered a second scenario, referred to as 57W-43PV, where the installed solar PV capacity is assumed to increase to 8.6 GW, while wind capacity rises to 11.45 GW. These values are based on targets outlined in the roadmap published by the 2024 Climate Action Plan~\citep{cap2024future}. This study does not include offshore wind in the analysis. Recent reports suggest that even by 2030, Ireland is unlikely to have any significant new offshore wind farms, with projected offshore capacity expected to remain near zero using realistic scenarios~\citep{seai2024future}.

@@

The accuracy of the datasets used in this study was verified against national generation data, before continuing to the analysis of RES droughts. For the verification process, time varying values of installed capacity were used to account for changes in RES development over the verification period. This validation step evaluates how well the datasets represent actual renewable generation profiles by comparing them against observed data. The overall statistical distribution of CF values for wind (2014–2023) and solar PV (2023) is presented in the violin plots in Fig.~\ref{fig:violin_plots}. These plots illustrate the density of CF values for each dataset, highlighting their differences and alignment with observations. The results indicate that ATL aligns more closely with observations for wind, while all datasets exhibit similar distributions for solar PV.

@@

Verification of wind generation data highlights the importance of selecting a representative wind turbine power curve for the region being analysed. The ATL dataset, which uses a representative wind turbine power curve, is skilled at reproducing wind CF and RES droughts across Ireland. On the other hand, the power curve used for both C3S~GRD and C3S~NAT is not representative for Ireland, as it severely overestimates generation, underestimating the occurrence of RES droughts. This highlights a problem with using generalised datasets for analysing RES droughts: biases severely affect their ability to accurately reproduce RES drought events. The skill scores for the three datasets (Table~\ref{tab:wind_skill_scores}) show only a small difference in their ability to reproduce the changes in CF, as seen by their similar CC scores. However, their ability to reproduce the actual CF values is much lower than that for ATL, with RMSE scores almost four times higher than for the two C3S datasets. There is a clear bias towards an overestimation of CF, seen in the MBE values, which leads to an underestimation of RES droughts, which highlights the strong motivation to use regionally verified models to assess RES droughts.

@@

In this section, RES droughts are analysed by calculating the frequency and duration of events, their return periods, and their seasonality. By examining both the frequency of occurrence and duration of these events, TSOs can better assess the need for alternative fossil fuel-based generation, long-duration energy storage, demand-side measures, and enhanced interconnections with neighbouring power systems to ensure adequate supply in a cost-effective manner. Results are presented for the three datasets, which clearly illustrate how different modelling assumptions influence the characterisation of RES droughts.

@@

RES drought events identified over the 45-year period were used to calculate the return periods for different RES drought durations. A return period is the estimated average time interval between events of a specified duration (not to be confused with the frequency of their occurrence within a fixed time frame). Fig.~\ref{fig:return_periods} shows the return periods for different RES drought durations, which can be used to capture the most extreme events affecting the system. Understanding their return periods is crucial to inform decision makers about the trade-off between the risk of extreme events and the associated costs of mitigation measures. Extreme but rare RES droughts pose the toughest challenge to energy security by placing significant strain on the conventional backup sources necessary to maintain security of supply during these events.

@@

Across Fig.~\ref{fig:return_periods}a, c, and d, the return periods in the ATL dataset are consistently higher than those in the two C3S datasets. For instance, in the 91W-9PV scenario (Fig.~\ref{fig:return_periods}c), an event with a one-year return period lasts six days in the ATL dataset, compared to only five days in the C3S datasets. This difference underscores the importance of dataset selection when quantifying RES droughts, as each dataset’s assumptions and parametrisations significantly influence RES drought duration estimates. Additionally, in all four graphs, the similarity between results from the two C3S datasets suggests that assumptions in the ATL dataset, such as wind turbine power curve selection and PV panel specifications, have a greater impact on RES drought duration estimates than the precise geographic distribution of RES farms when studying the return periods of RES droughts.

@@

The seasonal analysis of RES droughts is based on the percentage of hours in each month classified as part of a RES drought event. Wind droughts tend to be more frequent during summer, whereas solar PV droughts are more common in winter due to reduced sunlight. By comparing these seasonal patterns across different datasets and energy scenarios, this study examines how differences in model assumptions and capacity mix influence the characterisation of RES drought events. In Northwestern Europe, winter droughts are primarily driven by variations in wind generation, whereas summer droughts are mainly related to solar generation. This seasonal analysis is critical to ensure a cost-effective balance between generation and demand throughout the year.

@@

The effect of adding solar PV capacity to a wind dominated power system, as is currently happening in Northwestern Europe, has been explored in the context of RES droughts. Our analysis has demonstrated that transitioning to a system with more equal amounts of wind and solar PV capacity reduces the occurrence of RES drought events, mitigates extreme RES drought conditions and enhances overall system resilience. This improvement is attributed to the complementary nature of wind and solar PV generation, as solar PV generation typically peaks in summer while wind generation predominates during winter. However, this integration is unable to counter critical winter RES droughts, which coincide with the strongest electricity demand in Northern European countries.

The results presented in this study have three main limitations. First, the definition of RES droughts based on generation does not consider the important role of demand, which could be of interest to system operators. Second, recent solar PV capacity expansions have changed the generation profile, limiting solar PV data for model training to a single year, although a longer validation period would be preferable. Third, the source for weather data is ERA5 which has limited spatial resolution, an issue that can be addressed once higher resolution datasets become available.
